\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-24 chapters/gaze_independence.tex
All review comments

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010}
stimulation do not rely on gaze redirection but also result in lower \acp{itr}
compared to visual paradigms, with \acp{itr} ranging between 1 and 10 bits/min
for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.
While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent
variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

To summarize, for individuals with \ac{sspgi}, gazing directly at a screen target may
be uncomfortable, impractical, or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while \acp{bci} hold great promise for these individuals, conventional
gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
control.
Therefore, the development of decoding strategies that account for covert
\ac{vsa} becomes crucial in the pursuit of high-performance gaze-independent
\acp{bci}.

@@

This leads us to adopt a specific approach: \emph{improve the performance of a
spatially organized visual oddball \ac{erp}-based brain
computer interface by using a suited decoding strategy}.

To achieve our goal, we chose a suitable interface and innovate on decoder
development.
We collect data from healthy control subjects to test our decoders.
Finally, the findings of these experiments will be verified in experiments with
individuals with \ac{sspgi}, including individuals with \ac{lis}, evaluating
both performance and comfort.

This specific target is embedded within a broader goal of enabling
effective communication for those with \ac{sspgi}.
Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities.
We believe this can be done with a non-invasive, spatially
organized visual \acp{bci} with high \ac{itr}.
Finally, our efforts also aim to improve \ac{erp} decoding performance
in general, which will also contribute to their effectiveness in both gaze-dependent
and gaze-independent settings.

>>>>

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010} stimulation do not rely on gaze redirection
but also result in lower \acp{itr} compared to visual paradigms.
A survey in by \textcite{Riccio} from 2012 reported that \acp{itr} ranged between
1 and 10 bits/min for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.

On the tactile front, recent studies use steady-state somatosensory evoked
potentials~\cite{Petit2021} or tactile oddball paradigms~\cite{Waal2012}
These are analogous to the \ac{ssvep} and visuall oddball paradigms,
but with vibrotactile stimulation at different spatial locations and/or
frequencies~\cite{Han2020}.
Some noteworthy advancements achieved an \ac{itr} of 14.77
bits/min~\cite{Liu2019} and 20.73 bits/min~\cite{Herweg2016} with tactile
oddball stimulation through highly optimized stimulation?
Still, other recent tactile paradigm studies~\cite{Herweg2016a,Han2020,Jin2020,Li2019,Eidel2022}
fail to exceed 5 bits/min.

Auditory \acp{bci} are similarly implemented as paradigms oddball or
steady-state stimulation with spatial or tone modulation.
\textcite{Zhang2020} report that recent decoding methods can reach 16.99
bits/min \ac{itr}, but, in general, auditory \acp{bci} are also limited in
their speed and usually outperformed by visual paradigms~\cite{Rezeika2018}.

While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
settings when coupled with visual oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Auditory and tactile stimulation can even be paired to create a fully
non-visual, hybrid \ac{bci}, with a reported \ac{itr} of 11.66 bits/min~\cite{Zhang2020}.

Yet, auditory and somatosensory \acp{bci} also suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.
However, non-visual stimulation modalities are still valuable for developing
BCIs for individuals affected with severe vision loss or
blindness~\cite{Nijboer2008,Rutkowski2015}, which can occur in some cases of
\ac{lis}~\cite{Patterson1986}.

@@

To summarize, for individuals with \ac{sspgi}, gazing directly at a screen target may
be uncomfortable, impractical, or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while visual \acp{bci} hold great promise for these individuals,
if they do not suffer from severe vision loss, conventional
gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
control.
Therefore, the development of decoding strategies that account for covert
\ac{vsa} becomes crucial in the pursuit of high-performance gaze-independent
\acp{bci}.

@@

This leads us to adopt the following hypothesis: \emph{it is possible to improve
the (gaze-independent) accuracy of a spatially organized visual oddball
\ac{erp}-based brain computer interface by using a suited decoding strategy}.

To test this hypothesis, we chose a suitable interface, innovate on decoder
development and test if these innovations improve \ac{bci} performance in a
meaningful way.
Accordingly, we collected data from healthy control subjects to test proposed
decoders, and the findings have be verified in experiments with
individuals with \ac{sspgi}, including individuals with \ac{lis}, focusing
both accuracy and comfort.

The specific target of improving \ac{bci} performance under these conditions
is embedded within a broader goal of enabling effective communication for those with \ac{sspgi}.
Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities.
We believe this can be done with a non-invasive, spatially
organized visual \acp{bci} with high \ac{itr}.
Finally, our efforts also aim to improve \ac{erp} decoding performance
in general, which will also contribute to their effectiveness in both gaze-dependent
and gaze-independent settings.

### 2024-11-24 chapters/patients.tex
All review comments

\chapter{Case studies}
\label{sec:patients}
\epigraph{I knew there would be a lot of unexpected issues involved with
patient experiments, but I never thought there would be a baby
goats in the experiment room.}{Juliette}

@@

\section{Materials \& methods}
\subsection{Recruitment}
Participants were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France), and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must:
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according to the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they:
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
  \item be unable to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

@@

\section{Outcomes}

\subsection{Eye tracking analysis}
\label{sec:patients/outcomes/gaze}

First, we aimed to shed more light on the actual capabilities of individuals
with \ac{sspgi} regarding performing overt \ac{vsa} and central gaze
fixation, as well as to investigate how relevant these two settings are when the
gaze is not cued.
\Cref{fig:patients/gaze} maps gaze position relative to the stimuli
across conditions.
These results should be interpreted with care, as the eye tracker to some
degree relies on functioning eye motility.
The participant's position relative to the eye tracker might have shifted
throughout the experimental session despite our best efforts, e.g., because they needed aspiration of their
tracheostomy.

@@

Due to the heterogeneous nature of the participants' conditions, it is difficult to
draw general conclusions.
This study should therefore be seen as a collection of case studies,
highlighting different obstacles encountered in developing gaze-independent
visual oddball \acp{bci} for individuals with \ac{sspgi}.
Nevertheless, we would like to highlight some aspects that might be
of interest for the further development of this class of \acp{bci}.

@@

\section{Limitations}

@@

In this study, user comfort in the different conditions was not objectively
measured.
Instead, it was assumed that participants operated most comfortably in the free
\ac{vsa} condition.
To properly contextualize performance results, they should be coupled
with a measure of user comfort and perceived effort, like the NASA Task Load
Index~\cite{Hart2006}.
Performance might, after all, be traded off for user comfort.
Eye motor disability could also have been assessed more
objectively~\cite{FriedOken2020}, using, e.g.,
the Revised Coma Recovery Scale~\cite{Giacino2004} or the NSUCO
oculomotor~\cite{Maples1992}.

>>>>

\chapter{Case studies}
\label{sec:patients}
\epigraph{I knew there would be a lot of unexpected issues involved with
patient experiments, but I would never have guessed that there would be baby
goats in the experiment room\ldots}{Juliette}

@@

\section{Methods}
\subsection{Recruitment}
Participants were recruited across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France), and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must:
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according to the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they:
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
  \item be unable to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

@@

\section{Results}

\subsection{Visual skill and eye tracking analysis}
\label{sec:patients/outcomes/gaze}

Given these information, we aimed to shed more light on the actual capabilities of individuals
with \ac{sspgi} regarding performing overt \ac{vsa} and central gaze
fixation, as well as to investigate how relevant these two settings are when the
gaze is not cued.
\Cref{fig:patients/gaze} maps gaze position relative to the stimuli
across conditions.
These results should be interpreted with care, as the eye tracker to some
degree relies on functioning eye motility.
The participant's position relative to the eye tracker might have shifted
throughout the experimental session despite our best efforts, e.g., because they needed aspiration of their
tracheostomy.

@@

\subsection{Gaze-independent operation \& decoding}
Due to the heterogeneous nature of the participants' conditions, it is difficult to
draw general conclusions.
This study should therefore be seen as a collection of case studies,
highlighting different obstacles encountered in developing gaze-independent
visual oddball \acp{bci} for individuals with \ac{sspgi}.
Nevertheless, we would like to highlight some aspects that might be
of interest for the further development of this class of \acp{bci}.

@@

\subsection{Limitations}

@@

In this study, user comfort in the different conditions was not objectively
measured.
Instead, it was assumed that participants operated most comfortably in the free
\ac{vsa} condition.
To properly contextualize performance results, they should be coupled with
metrics evaluating the user's requirements
with a measures of usability, comfort and perceived effort, like the NASA Task Load
Index~\cite{Hart2006} and other metrics proposed in the user-centered design
framework for \acp{bci}~\cite{Kuebler2014}.
Performance might, after all, be traded off for user comfort.
Eye motor disability could also have been assessed more
objectively~\cite{FriedOken2020}, using, e.g.,
the Revised Coma Recovery Scale~\cite{Giacino2004} or the NSUCO
oculomotor exam~\cite{Maples1992}.

### 2024-11-24 chapters/stbf_struct.tex
All review comments

	\subsubsection{Riemannian geometry classifier}
	\label{sec:riemannian}
	We opted for a Riemannian
	geometry-based classifier to compare our results.
	The Riemannian model (\textsc{xdawn+rg}) uses the XDAWN spatial filter combined
	with Riemannian geometry in tangent space as implemented by Barachant et
	al.~\cite{Barachant2014a}.
	This classifier uses four XDAWN spatial filters and each epoch's empirical spatial covariance matrix.
	The target with the maximum score is the prediction of the cued target.
	\textsc{xdawn+rg} was trained and validated without downsampling using epochs
	at the original sample rate of 2048 Hz.

@@

	\subsection{Classifier training time}
	In order to evaluate the training time of the investigated classifiers, the
	cross-validation scheme is run four times for each subject, each time with an
	increasing number of \ac{eeg} channels retained in the analysis, to explore the scalability of each classifier for analyses with higher spatial resolutions.
	The temporal resolutions were not varied, but we expect that increasing the
	temporal resolution has a similar effect on training time since the
	training times for the \textsc{stbf}-based as evidenced by the complexity
	calculations in \cref{sec:stbf-struct/discussion/param-complex}.
	\cref{fig:stbf-struct/training-time} shows the measured training times.
	These results were obtained using a laptop with an Intel® Core™ i7-8750H CPU and 16GB of RAM.

	\begin{figure}
		%\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
    \sffamily
    \sansmath
    \input{figures/stbf_struct/training_time.pgf}
    \caption[Classifier training time.]{Training time boxplots of different
    classifiers and individual results for all training folds for all subjects
    using 32 channels. Using structured covariance estimation greatly reduces
    training time.}
		\label{fig:stbf-struct/training-time}
	\end{figure}

	\Cref{fig:stbf-struct/training-time} shows that the training time of
  \textsc{stbf-struct} has a lower median training time (0.06 s) than
  \textsc{stbf-shrunk} (0.33 s), \textsc{stbf-emp} (0.30 s) and
  \textsc{xdawn+rg} (1.04 s) when training
  on all available data and using 32 channels.

@@

	The results shown in \cref{fig:stbf-struct/training-time} also confirm that the
	\ac{stbf}-based estimators are very fast to train compared to the
	state-of-the-art estimator \textsc{xdawn+rg}, which confirms the results
  reported by~\textcite{Wittevrongel2016}.
	Since the training times of all \ac{stbf}-based classifiers are already in
	the order of tenths of seconds, the question arises whether the
	improvements achieved by using the structured estimator would be relevant in
	practice.
	However, the authors believe that these results could significantly impact some
	use cases of the spatiotemporal beamformer, like high spatial or temporal
  resolution \ac{erp} analyses (e.g., $C>>32, S>>256$).
	One example is single-trial \ac{erp} analysis with a high-temporal
	resolution to extract \ac{erp} time features.
	Such higher-resolution analyses can later be incorporated into an \ac{erp}
	classification framework.
	In addition, the speed-up provided by structured estimation yields a faster
	off-line evaluation of the \ac{stbf} \ac{erp} classifier, where often multiple cross-validation folds, subjects, and hyperparameter settings need to be explored, which can quickly increase runtime.
	Improvements in computation speed and memory usage can remove the need for dedicated computation hardware and enable running group analyses on a personal computer.

>>>>

	\subsubsection{Riemannian geometry classifier}
	\label{sec:riemannian}
	We opted for a Riemannian
	geometry-based classifier to compare our results.
	The Riemannian model (\textsc{xdawn+rg}) uses the XDAWN spatial filter combined
	with Riemannian geometry in tangent space as implemented by Barachant et
	al.~\cite{Barachant2014a}.
	This classifier uses four XDAWN spatial filters and each epoch's empirical spatial covariance matrix.
	The target with the maximum score is the prediction of the cued target.
	\textsc{xdawn+rg} was trained and validated without downsampling using epochs
	at the original sample rate of 2048 Hz.
  Preliminary experimentation revealed that best performance in low and high
  data availability setting was achieved when no shrinkage was applied to the estimation of XDAWN filters or the subsequent
  covariance matrices for this specific dataset.
  Hence, no shrinkage regularization will be applied to \textsc{xdawn+rg} in
  further experiments.

@@

	\subsection{Classifier training time}
	In order to evaluate the training time of the investigated classifiers, the
	cross-validation scheme is run for each subject.
  For this analysis, downsampling to 32 Hz was also performed for the XDAWN+RG
  classifier for fair comparison.
  \cref{fig:stbf-struct/training-time} shows the measured training times.
	These results were obtained using a laptop with an Intel® Core™ i7-8750H CPU and 16GB of RAM.

	\begin{figure}
		%\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
    \sffamily
    \sansmath
    \input{figures/stbf_struct/training_time.pgf}
    \caption[Classifier training time.]{Training time boxplots of different
    classifiers and individual results for all training folds for all subjects
    using 32 channels and a sampling rate of 32 Hz. Using structured covariance estimation greatly reduces
    training time.}
		\label{fig:stbf-struct/training-time}
	\end{figure}

	\Cref{fig:stbf-struct/training-time} shows that the training time of
  \textsc{stbf-struct} has a lower median training time (0.06 s) than
  \textsc{stbf-shrunk} (0.33 s), \textsc{stbf-emp} (0.30 s) and
  \textsc{xdawn+rg} (0.20 s) when training
  on all available data and using 32 channels.

@@

	Since the training times of all \ac{stbf}-based classifiers are already in
	the order of tenths of seconds, the question arises whether the
	improvements achieved by using the structured estimator would be relevant in
	practice.
	However, the authors believe that these results could significantly impact some
	use cases of the spatiotemporal beamformer, like high spatial or temporal
  resolution \ac{erp} analyses (e.g., $C>>32, S>>256$).
	One example is single-trial \ac{erp} analysis with a high-temporal
	resolution to extract \ac{erp} time features.
	Such higher-resolution analyses can later be incorporated into an \ac{erp}
	classification framework.
	In addition, the speed-up provided by structured estimation yields a faster
	off-line evaluation of the \ac{stbf} \ac{erp} classifier, where often multiple cross-validation folds, subjects, and hyperparameter settings need to be explored, which can quickly increase runtime.
	Improvements in computation speed and memory usage can remove the need for dedicated computation hardware and enable running group analyses on a personal computer.
